---
title: "geos2916_hydrology"
author: "Tracy"
format: html
---

# Importing data

```{r}
# libraries
library(tidyverse)
library(ggplot2)
library(readxl)
library(janitor)
library(patchwork)
library(ggpmisc)

# plotting
theme_set(theme_classic())
palette = (c("orange", "lightgreen",  "lightblue"))
```

```{r message=FALSE}
# importing data
flow <- read_excel("data/2024_DischargeDataWeek8Excel.xlsx",
                   range="A2:R1097", col_names=TRUE) |> 
        select(-c(6,12,13)) # remove empty columns
```

```{r}
# missing data check
# visdat::vis_miss(flow)
```

```{r}
# separating stations
wagga <- flow |> select(1:5)
gundagai <- flow |> select(6:10)
narrandera <- flow |> select(11:15)

# renaming columns
newcolnames=c("day", "month", "year", "flow_ML_day", "date")

colnames(wagga)=newcolnames
colnames(gundagai)=newcolnames
colnames(narrandera)=newcolnames

# month abbreviations

wagga <- wagga
gundagai <- gundagai
narrandera <- narrandera

# final data check
wagga; gundagai; narrandera
```

# Q1: Daily stream power

Where stream power is calculated using:

> $Ω = pgQS$

```{r}
# constants
p=1000 # density of water (kg/m^3)
g=9.8 # gravitational constant (m/s^2)

# streamflow conversion
ML_m3=1000 # cubic metres in 1 megalitre
t=86400 # length of a day (s)

# local valley slopes (m/m)
S_w=3.2*10^3 # wagga wagga
S_g=0.16 # gundagai
S_n=9.2*10^3 # narrandera

# stream power (kg/m^2/s^3)
stream_power <- function(station, slope) {
  
  # daily stream power
  station |> filter(between(year, 2009, 2011)) |>
    mutate(
      q = flow_ML_day*ML_m3/t, # si unit conversion, flow (m^3/s)
      streampower = p*g*q*slope) |>
    
  # monthly average
  group_by(month) |>
    summarise(
      log_streampower = mean(log10(streampower))) |>
    mutate(
      month_abb=month.abb[month],
      month=NULL # no longer needed
    )
}

wagga_monthly <- stream_power(wagga, S_w)
gundagai_monthly <- stream_power(gundagai, S_g)
narrandera_monthly <- stream_power(narrandera, S_n)

# final check
wagga_monthly; gundagai_monthly; narrandera_monthly
```

# Q2: Monthly mean stream flow

## a. Calculations

```{r}
# monthly naming + ordering
month <- function(station) {
  factor(station$month_abb, levels=unique(station$month_abb))}

# ensures months keep order when plotted
wagga_monthly$month_abb=month(wagga_monthly)
gundagai_monthly$month_abb=month(gundagai_monthly)
narrandera_monthly$month_abb=month(narrandera_monthly)

# final check
wagga_monthly; gundagai_monthly; narrandera_monthly
```

```{r}
# combined dataset for plotting
wagga_monthly$station="Wagga Wagga"
gundagai_monthly$station="Gundagai"
narrandera_monthly$station="Narrandera"

combined <- rbind(wagga_monthly, gundagai_monthly, narrandera_monthly) |> mutate(month=month_abb, month_abb=NULL) |>
  select("month", "station", "log_streampower")

combined
```


## b. Plotting

```{r}
# combined
log_monthly_sp <- ggplot(combined, aes(fill=station, y=log_streampower, x=month)) + 
    geom_bar(position="dodge", stat="identity") +
  labs(x="Month", fill="Station",
       y=bquote(Log~mean~stream~power~(kg~m^-2~s^-3)))

ggsave("figures/log_monthly_sp.png", height=4, plot=log_monthly_sp)
```

```{r}
# individual
graph_steampower <- function(data) {
data |> ggplot(aes(x=month_abb, y=10^log_streampower)) + 
  labs(x="Month", y=bquote(Mean~Monthly~Stream~Power~(kg~m^-2~s^-3)))
}

wagga_plot <- graph_steampower(wagga_monthly) + geom_col(col="gold", fill="gold")
gundagai_plot <- graph_steampower(gundagai_monthly) + geom_col(col="blue", fill="blue")
narrandera_plot <- graph_steampower(narrandera_monthly) + geom_col(col="red", fill="red")

# saving plots
ggsave("figures/wagga_flow.png", height=4, plot=wagga_plot)
ggsave("figures/gundagai_flow.png", height=4, plot=gundagai_plot)
ggsave("figures/narrandera_flow.png", height=4, plot=narrandera_plot)
```

# Q3: Monthly specific stream power

## a. Calculations

Specific stream power (w) is obtained using:

> $w =\frac{Ω}{b}$

```{r}
# bankfull stream width estimates (m)
b_w=82
b_g=70
b_n=92

# mean monthly "specific" stream power
specific_sp <- function(data_monthly, bankfull_width) {
  data_monthly |>
    mutate(log_specific_sp = log10((10^log_streampower)/bankfull_width)) |>
    subset(select=c("month_abb", "log_specific_sp"))
}

wagga_monthly_sp <- specific_sp(wagga_monthly, b_w)
gundagai_monthly_sp <- specific_sp(gundagai_monthly, b_g)
narrandera_monthly_sp <- specific_sp(narrandera_monthly, b_n)

# final check
wagga_monthly_sp; gundagai_monthly_sp; narrandera_monthly_sp
```

```{r}
# combined dataset for plotting
wagga_monthly_sp$station="Wagga Wagga"
gundagai_monthly_sp$station="Gundagai"
narrandera_monthly_sp$station="Narrandera"

combined_sp <- rbind(wagga_monthly_sp, gundagai_monthly_sp, narrandera_monthly_sp) |> mutate(month=month_abb, month_abb=NULL) |>
  select("month", "station", "log_specific_sp")

combined_sp
```

## b. Plotting

```{r}
# combined
log_monthly_specific_sp <- ggplot(combined_sp, aes(fill=station, y=log_specific_sp, x=month)) + 
    geom_bar(position="dodge", stat="identity") +
  labs(x="Month", fill="Station",
       y=bquote(Log~mean~specific~stream~power~(kg~m^-2~s^-3)))

ggsave("figures/log_monthly_specific_sp.png", height=4, plot=log_monthly_specific_sp)
```

```{r}
# individual
graph_steampower <- function(data) {
data |> ggplot(aes(x=month_abb, y=10^log_specific_sp)) + 
  labs(x="Month", y=bquote(Mean~Monthly~Specific~Stream~Power~(kg~m^-2~s^-3)))
}

wagga_sp_plot <- graph_steampower(wagga_monthly_sp) + geom_col(col="gold", fill="gold")
gundagai_sp_plot <- graph_steampower(gundagai_monthly_sp) + geom_col(col="blue", fill="blue")
narrandera_sp_plot <- graph_steampower(narrandera_monthly_sp) + geom_col(col="red", fill="red")

# saving plots
ggsave("figures/wagga_sp_flow.png", height=4, plot=wagga_sp_plot)
ggsave("figures/gundagai_sp_flow.png", height=4, plot=gundagai_sp_plot)
ggsave("figures/narrandera_sp_flow.png", height=4, plot=narrandera_sp_plot)

wagga_sp_plot/gundagai_sp_plot/narrandera_sp_plot
```

# Q4: Murrumbidgee flooding

## a. Sediment flux calculations

The minimum energy slope can be calculated using:

> $S_{f\_min}=(\frac{Q_s}{Q})^{0.708}*Q^{-0.080}$

```{r}
# flow (10^6 L/day)
flow_w=32326
flow_g=32637

# mean sediment concentration (10^-3 g/L)
sc_w=227
sc_g=50.8

# sediment flux (T/day -> 10^6 g/day)
sediment_flux <- function(flow, concentration) {concentration*flow/10^6} # tonnes/day

sf_w <- sediment_flux(flow_w, sc_w)
sf_g <- sediment_flux(flow_g, sc_g)

# peak daily discharge
peakd_w <- wagga |> filter(year==2011) |>
  mutate(q=flow_ML_day*ML_m3/t) |> select(q) |> max()
peakd_g <- gundagai |> filter(year==2011) |>
  mutate(q=flow_ML_day*ML_m3/t) |> select(q) |> max()

# minimum energy slope (m/m)
s_minimum <- function(sediment_flux, discharge) {
  (sediment_flux/discharge)^0.708*(discharge^-0.080)}

s_fmin_w <- s_minimum(sf_w, peakd_w) 
s_fmin_g <- s_minimum(sf_g, peakd_g)
```

## b. Minimum energy slope vs local valley slope

```{r}

```
